\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{margin=2.5cm}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{enumitem}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!10},
}

\title{Knowledge Graph Completion Analysis}
\author{Lucas Perez Reis Lobo, Yutaka Shibata, Sofia Davis, \\ William Caulton, Mert Yerlikaya, Davyd Shtepa \\ \small Knowledge Engineering CW2 --- Music History \& Heritage KG}
\date{April 2026}

\begin{document}
\maketitle

\section{Overview}

This document analyses the completeness of our Music History knowledge graph (42,554 triples, 58 primary artists) and describes the strategy and results of using RAG to resolve identified gaps.

We apply the completeness metrics from Zaveri et al.\ (2015), as taught in Week 10: \textbf{CM1} (schema completeness), \textbf{CM2} (property completeness), \textbf{CM3} (population completeness), and \textbf{CM4} (interlinking completeness).

The completion analysis was discussed with the team after the third group meeting. The domain experts confirmed that the 5 ontology and 5 instance gaps below accurately reflect the most impactful incompleteness in the KG, and agreed that RAG should be applied to all 10 gaps.

%--------------------------------------------------------------
\section{5 Incomplete Ontology Elements}
%--------------------------------------------------------------

Table~\ref{tab:ontology-gaps} summarises the 5 ontology-level gaps, their completeness metrics, and the RAG-driven resolution for each.

\begin{table}[h]
\centering
\small
\begin{tabular}{p{0.8cm}p{3.5cm}p{2cm}p{3cm}p{3cm}}
\toprule
\textbf{ID} & \textbf{Gap} & \textbf{Metric} & \textbf{How to complete} & \textbf{RAG result} \\
\midrule
O1 & \texttt{signedTo} has no temporal scoping. CQ17 produces cartesian products (229 triples). & CM1: no start/end dates & RDF reification or named graphs. Wikidata P264 qualifiers. & P34: 226 temporal ranges (46\% high conf.). Not implemented. \\[6pt]
O2 & \texttt{performedAt} had only 5 triples (3 artists). No structured concert data. & CM1: minimal coverage & Setlist.fm API or MusicBrainz events. & P35: 113 events. \textbf{Final: 131 triples} (26$\times$). \\[6pt]
O3 & \texttt{collaboratedWith} modelled via band entities, not direct artist links. 76 triples. & CM1: low density & LLM text extraction for direct collaborations. & P36: 69 new pairs. \textbf{Final: 242 triples} (3.2$\times$). \\[6pt]
O4 & \texttt{founded} had only 4 triples. Artists founding bands, labels, orgs not captured. & CM2: $<$1\% & Extend LLM prompt. Wikidata P112 (founded by). & P37: 47 new entries. \textbf{Final: 69 triples} (17$\times$). \\[6pt]
O5 & \texttt{hasMusicalPeriod} had 5 triples (3 artists). No period classification. & CM1: minimal & Wikidata P2348 or LLM classification. & P38: 112 assignments. \textbf{Final: 126 triples} (all 58 artists). \\
\bottomrule
\end{tabular}
\caption{5 incomplete ontology elements with completeness metrics and RAG resolution.}
\label{tab:ontology-gaps}
\end{table}

\textbf{O1} is the only unresolved ontology gap. The \texttt{signedTo} property links each artist to all their record labels simultaneously without temporal context. RAG prompt P34 demonstrated that an LLM can provide approximate year ranges (226 pairs, 46\% high confidence), but implementing temporal scoping requires RDF reification or named graphs, which would add significant modelling complexity. The team agreed to document this as future work.

\textbf{O2--O5} were resolved through a combination of RAG prompts and automated LLM extraction. The most dramatic improvement was \texttt{performedAt} (5 $\rightarrow$ 131 triples, a 26$\times$ increase), followed by \texttt{founded} (4 $\rightarrow$ 69, 17$\times$) and \texttt{hasMusicalPeriod} (5 $\rightarrow$ 126, now covering all 58 primary artists).

%--------------------------------------------------------------
\section{5 Incomplete Instance Elements}
%--------------------------------------------------------------

Table~\ref{tab:instance-gaps} summarises the 5 instance-level gaps. Unlike the ontology gaps, several of these were fully resolved through pipeline improvements.

\begin{table}[h]
\centering
\small
\begin{tabular}{p{0.8cm}p{3.5cm}p{2cm}p{3cm}p{3cm}}
\toprule
\textbf{ID} & \textbf{Gap} & \textbf{Metric} & \textbf{Resolution approach} & \textbf{Result} \\
\midrule
I1 & 1,274 secondary artists lack genre, country, dates. Only 58 primary are fully enriched. & CM2: genre 4.6\%, country 11.4\% & RAG sample (P39) + pipeline extension. & P39: 50 enriched (94\%). Pipeline: +216 triples for 102 artists. \\[6pt]
I2 & Umm Kulthum matched wrong MB entity (``UMM''). 4 wrong triples. Cascading failure. & CM3: severely under-represented & RAG (P40) + programmatic MBID search + config override. & \textbf{Fully resolved.} 58/58 artists. LLM hallucinated MBID. \\[6pt]
I3 & LLM text extraction: 16/58 artists (28\%). 42 missing text-unique data. & CM3: 28\% coverage & RAG demo (P41) + Gemini API automation. & \textbf{58/58 (100\%)}. 897 triples. 44\% text-unique predicates. \\[6pt]
I4 & Hildegard von Bingen: 0 genres. Only primary artist without genre. & CM2: 1/58 missing & RAG (P42) for pre-modern music. & \textbf{Fully resolved.} 5 genres, 3 periods, 10 compositions. \\[6pt]
I5 & 330 cover composers lack country, genre, dates. Only names and MBIDs. & CM2: uncharacterised & RAG sample (P43) + enrichment extension. & P43: 50 enriched (94\%). Pipeline extended for auto-enrichment. \\
\bottomrule
\end{tabular}
\caption{5 incomplete instance elements with completeness metrics and resolution results.}
\label{tab:instance-gaps}
\end{table}

\textbf{I2 (Umm Kulthum)} was the most instructive gap. The pipeline's MusicBrainz search matched ``UMM'' (an electronic project) instead of the Egyptian singer because the correct entry uses the Arabic name as primary. This single entity resolution error cascaded through the entire pipeline: wrong MB entity $\rightarrow$ no Wikidata cross-reference $\rightarrow$ no Wikipedia text $\rightarrow$ no LLM extraction $\rightarrow$ no genres, awards, or relationships. RAG prompt P40 provided comprehensive data but hallucinated the MBID (returned HTTP 404). The correct MBID was found via programmatic MusicBrainz API search, confirming that LLMs are unreliable for external identifiers. The fix was a config-driven MBID override, and all 4 sources now fetch correctly.

\textbf{I3 (text extraction)} progressed from a RAG demonstration to full pipeline automation. P41 extracted 87 triples for 5 diverse artists, finding that 44\% use predicates no structured API provides (\texttt{alterEgo}, \texttt{producedBy}, \texttt{pioneerOf}). The same prompt template was then integrated via Google Gemini free-tier API, raising coverage from 16/58 (28\%) to 58/58 (100\%) with 897 total text-extracted triples.

\textbf{I4 (Hildegard)} is the clearest case of RAG compensating for structured source blind spots. MusicBrainz, Wikidata, and Discogs all failed to provide genre data because their taxonomies do not cover pre-modern music. The LLM provided appropriate musicological terms (plainchant, monophonic music, sacred music) that no automated source could.

\textbf{I1 and I5} were addressed through both RAG sampling (50 entities each, 94\% resolved) and pipeline extensions. The \texttt{enrich\_related\_artists} function was extended to fetch genres, birth/death dates, and type (not just country), and the enrichment query was broadened to include artists linked via \texttt{influencedBy} and \texttt{composed}.

%--------------------------------------------------------------
\section{RAG Strategy and Results}
%--------------------------------------------------------------

\subsection{Approach}

We use RAG (Week 11) to resolve incompleteness via a 6-step cycle:
\begin{enumerate}
  \item \textbf{Identify} a gap from the analysis above
  \item \textbf{Query} the KG via SPARQL to retrieve relevant context
  \item \textbf{Verbalise} retrieved triples into natural language
  \item \textbf{Prompt} an LLM with the context + gap question
  \item \textbf{Parse} the response into structured JSON
  \item \textbf{Ingest} into the KG via programmatic entity linking
\end{enumerate}

The ingestion script (\texttt{ingest\_rag\_results.py}) is integrated into the pipeline and runs automatically:

\begin{center}
\fbox{\textit{Identify gap $\rightarrow$ SPARQL retrieval $\rightarrow$ RAG prompt $\rightarrow$ JSON response $\rightarrow$ Programmatic ingestion $\rightarrow$ Enriched KG}}
\end{center}

\subsection{Example: Filling producer relationships}

To illustrate the RAG cycle concretely, we demonstrate gap-filling for Michael Jackson's producer relationships.

\textbf{Step 1. SPARQL retrieval:}
\begin{lstlisting}
PREFIX mh: <http://example.org/music-history/>
PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>
PREFIX dc: <http://purl.org/dc/elements/1.1/>
SELECT ?albumTitle ?releaseDate WHERE {
    ?artist rdfs:label "Michael Jackson"@en .
    ?artist mh:released ?album .
    ?album dc:title ?albumTitle .
    OPTIONAL { ?album mh:releaseDate ?releaseDate }
}
\end{lstlisting}
This returned 25 albums from the KG, including release dates. The existing KG already contained 3 \texttt{produced} triples linking Quincy Jones to Off the Wall, Thriller, and Bad (from the LLM text extraction phase).

\textbf{Step 2. Verbalisation:} The SPARQL results were converted to natural language: ``Michael Jackson released the following albums: Got to Be There (1972), Ben (1972), Off the Wall (1979), Thriller (1982)\ldots''

\textbf{Step 3. RAG prompt:} ``Based on what we know about Michael Jackson's discography, who produced each of these albums? Our KG already shows Quincy Jones for 3 albums. Context: [verbalised triples]. Return as JSON with confidence levels.''

\textbf{Step 4. LLM response and evaluation:}

\begin{table}[h]
\centering
\small
\begin{tabular}{lll}
\toprule
\textbf{Album} & \textbf{Producer (RAG)} & \textbf{Assessment} \\
\midrule
Got to Be There (1972) & Hal Davis & High confidence. Well-documented Motown production. \\
Ben (1972) & Hal Davis & High confidence. \\
Off the Wall (1979) & Quincy Jones & Confirmed $\checkmark$ (matches existing KG). \\
Thriller (1982) & Quincy Jones & Confirmed $\checkmark$ \\
Bad (1987) & Quincy Jones & Confirmed $\checkmark$ \\
Dangerous (1991) & MJ / Teddy Riley & Uncertain. LLM flagged multiple producers. \\
HIStory (1995) & Michael Jackson & Uncertain. Collaborative production. \\
Invincible (2001) & Rodney Jerkins & Uncertain. Fragmented production credits. \\
\bottomrule
\end{tabular}
\caption{RAG response for Michael Jackson producer relationships.}
\end{table}

The LLM confirmed 3 existing KG triples, provided 2 high-confidence new ones (Hal Davis), and proactively \textbf{flagged 3 as uncertain} where production was distributed across multiple collaborators. This demonstrates responsible KG construction: the LLM self-assesses confidence rather than asserting everything as fact. Only the 2 high-confidence triples were ingested; the 3 uncertain ones were documented for future manual validation.

\subsection{Results Summary}

\begin{table}[h]
\centering
\small
\begin{tabular}{lllr}
\toprule
\textbf{Gap} & \textbf{Prompt} & \textbf{RAG Output} & \textbf{Key Metric} \\
\midrule
O1 signedTo temporal & P34 & 226 temporal ranges & 46\% high confidence \\
O2 performedAt & P35 & 113 new events & 5 $\rightarrow$ 131 triples \\
O3 collaboratedWith & P36 & 69 new direct pairs & 76 $\rightarrow$ 242 triples \\
O4 founded & P37 & 47 new entries & 4 $\rightarrow$ 69 triples \\
O5 hasMusicalPeriod & P38 & 106 assignments & 5 $\rightarrow$ 126 triples \\
I1 secondary artists & P39 & 50 enriched & 94\% resolved \\
I2 Umm Kulthum & P40 & Full reconstruction & Entity fixed \\
I3 text extraction & P41 & 87 triples (5 demo) & 28\% $\rightarrow$ 100\% \\
I4 Hildegard genres & P42 & 5 genres, 3 periods & Genre-less fixed \\
I5 cover composers & P43 & 50 enriched & 94\% resolved \\
\bottomrule
\end{tabular}
\caption{10 RAG completion prompts (P34--P43) and their results. Total ingested: 494 triples.}
\end{table}

Beyond gap-filling, RAG served as a \textbf{data quality auditor} throughout the project. The KG vs Plain LLM comparisons (P18--P23) identified 6 actionable issues that were fed back into the pipeline:
\begin{itemize}
  \item ``interview'', ``acoustic'', ``ballad'', and ``classic rock'' were tagged as genres but are not musical genres. These were added to the genre blacklist.
  \item ``voice'' and ``lead vocals'' were separate instruments for the same concept. Instrument normalisation was added.
  \item ``eponymous'' appeared as an instrument due to a MusicBrainz tag artifact.
\end{itemize}

A key outcome of the RAG process was the \textbf{progression from manual demonstration to pipeline automation}. Several RAG findings were subsequently implemented as permanent pipeline improvements:
\begin{itemize}
  \item \textbf{I3}: The text extraction prompt from P41 was automated via Google Gemini API, increasing coverage from 16 to 58 artists.
  \item \textbf{I1/I5}: The enrichment function was extended to fetch genres and dates, not just country data.
  \item \textbf{Data quality}: Wikidata P31/P279* class hierarchy queries now programmatically filter for music genres (Q188451) and music awards (Q4220920), replacing the manual blacklists with a self-correcting approach.
\end{itemize}

%--------------------------------------------------------------
\section{Completeness Metrics Summary}
%--------------------------------------------------------------

\begin{table}[h]
\centering
\small
\begin{tabular}{llll}
\toprule
\textbf{Metric} & \textbf{Initial} & \textbf{Final} & \textbf{Detail} \\
\midrule
CM1 --- Schema & 93\% & 97\% & 29/30 object properties populated \\
CM2 --- Primary & High & Higher & Country 98\%, Genre 100\% \\
CM2 --- All artists & Low & Improved & +216 triples from extended enrichment \\
CM3 --- Population & 58 primary & Same & Data density increased \\
CM4 --- Interlinking & Strong & Stronger & Umm Kulthum cross-refs fixed \\
Text extraction & 16/58 (28\%) & 58/58 (100\%) & 897 triples via LLM API \\
Defined classes & 74 & 191 & +158\% \\
Total triples & $\sim$38,000 & 42,554 & +12\% \\
\bottomrule
\end{tabular}
\caption{Completeness metrics before and after all completion activities.}
\end{table}

%--------------------------------------------------------------
\section{Reflection}
%--------------------------------------------------------------

The completion process revealed several insights about automated KG construction, the role of LLMs, and the balance between automation and human oversight.

\subsection{RAG as a data quality auditor}

The most unexpected value of RAG was not gap-filling but \textbf{data quality auditing}. Every KG vs Plain LLM comparison (P18--P23) surfaced issues that were invisible to both the pipeline developer and automated validation. For example, P21 revealed that ``interview'' had leaked into the genre taxonomy from MusicBrainz tags, P20 found that ``voice'' and ``lead vocals'' were duplicated as separate instruments, and P22 identified that ``classic rock'' is a radio format, not a musical genre.

These findings triggered structural pipeline fixes (genre blacklist, instrument normalisation), which in turn improved the KG for all 58 artists, not just the ones tested. The iterative cycle of \textbf{build $\rightarrow$ evaluate with RAG $\rightarrow$ identify issues $\rightarrow$ fix pipeline $\rightarrow$ rebuild} proved more effective than either pure automation (which would leave the noise undetected) or pure manual curation (which would not scale to 42,554 triples).

\subsection{LLM limitations: external identifiers and non-Latin scripts}

The Umm Kulthum case (I2) exposed two significant LLM limitations.

First, \textbf{LLMs hallucinate external identifiers}. When RAG prompt P40 asked for the correct MusicBrainz MBID, the LLM returned \texttt{0c326960}, a plausible-looking UUID that returned HTTP 404 when queried. The actual correct MBID (\texttt{0a354a70}) was only found via programmatic MusicBrainz API search. This means external IDs (MBIDs, Wikidata QIDs, Discogs IDs) should never be taken from LLM responses without programmatic verification.

Second, \textbf{non-Latin scripts cause cascading failures}. The romanised name ``Umm Kulthum'' fuzzy-matched to ``UMM'' (an unrelated electronic project) because MusicBrainz stores the Arabic name as primary. This single entity resolution error propagated through the entire pipeline: wrong MB entity $\rightarrow$ no Wikidata cross-reference $\rightarrow$ no Wikipedia text $\rightarrow$ no LLM extraction $\rightarrow$ no genres, awards, or relationships. The artist node was effectively empty despite Umm Kulthum being one of the 20th century's most important musicians. The same issue affected Pyotr Ilyich Tchaikovsky (Cyrillic), which also required an MBID override.

This is a systemic limitation for any KG pipeline covering multilingual domains. Our solution (config-driven MBID overrides) is practical but not scalable. A production system would need either Wikidata-first entity resolution (searching by QID rather than name) or multilingual fuzzy matching across scripts.

\subsection{The automation-validation trade-off}

The pipeline automates KG construction end-to-end: data fetching from 4 APIs, structured and text mapping, entity linking, validation, RAG ingestion, and defined class assertion. However, the completion process showed that \textbf{three categories of human intervention remain essential}:

\begin{enumerate}
  \item \textbf{Ontological correctness.} The TTL file was loaded into Prot\'{e}g\'{e} to verify that OWL2 property characteristics (symmetric, transitive, irreflexive, asymmetric), inverse pairs, and class hierarchies matched the team's modelling decisions. The pipeline generates ontology declarations programmatically, but only visual inspection in Prot\'{e}g\'{e} can confirm that the semantics match the design intent. For example, we discovered that several properties lacked OWL2 characteristics (e.g., \texttt{covers} should be asymmetric and irreflexive) that were only identified through manual ontology review.

  \item \textbf{Semantic accuracy.} A domain expert review of the TTL identified 76 non-music genres (film/game categories from Wikidata), 31 non-music awards (military medals, honorary degrees), 19 range violations (LLM confusing ``released'' with ``composed''), and 10 duplicate entities. These were all \textit{structurally valid} RDF triples that were \textit{semantically wrong} in our domain. Automated validation catches structural errors (self-references, type mismatches), but domain knowledge is needed to catch semantic errors.

  \item \textbf{Multilingual coverage.} Name-based entity resolution fails for non-Latin scripts. This requires either manual MBID overrides or multilingual search capabilities that our current pipeline does not have.
\end{enumerate}

The optimal approach is a hybrid: automate construction, then iterate with human review. Our pipeline follows this cycle: \textbf{build $\rightarrow$ validate (automated) $\rightarrow$ review (Prot\'{e}g\'{e} + domain expert) $\rightarrow$ trace root causes $\rightarrow$ fix structurally $\rightarrow$ rebuild}. Each human-identified issue was traced to its root cause and fixed in the pipeline code, not patched in the TTL file. This ensures that the fixes persist across rebuilds and apply to all entities, not just the ones where the issue was spotted.

\subsection{Remaining Future Work}

\begin{enumerate}
  \item \textbf{Temporal scoping for \texttt{signedTo}}: Wikidata P264 qualifiers contain start/end dates for label contracts. Extracting these would require RDF reification or named graphs to qualify the existing \texttt{signedTo} triples, which would resolve the cartesian product issue in CQ17.
  \item \textbf{Concert/venue data}: Setlist.fm API or MusicBrainz events could provide structured performance data, reducing dependence on LLM extraction for \texttt{performedAt} triples.
  \item \textbf{Multilingual entity resolution}: Implementing cross-script fuzzy matching or adopting a Wikidata-first resolution strategy (searching by QID rather than romanised name) would eliminate the need for manual MBID overrides and make the pipeline robust for non-Western artists.
\end{enumerate}

\end{document}
